---
title: "Curating Data for Hypothesis Testing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Curating Data for Hypothesis Testing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Curating Data for Hypothesis Testing

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = F,
  comment = "#>"
)
```

```{r setup, echo=FALSE}
library(BioShiftR)
library(dplyr)
```


Here, we demonstrate how BioShifts and the BioShiftR package can be used to access, subset, and organize bioshifts data for research, hypothesis testing, and connecting to external data sources. 

## Example Scenario

Perhaps we want to evaluate the extent that bird ranges in Europe are shifting.

### Built-in filter steps 

Many of BioShiftR's functions provide simple methods for targeting, filtering, and subsetting range shift data. The `get_shifts()` function, for example, has arguments for continent, type (of range shift), and group (broad taxonomic and functional classifications); see the `get_shifts()` function page for details on options. 

Other functions which merge values from other dataframes to the range shifts database also have filtering options, usually for selecting study-level or species-specific values. For example, the `add_cv()`, `add_baselines()`, and `add_poly_info()` functions all contain options to either add these values from species-level or study-level polygons. 

### Get the data

In this case, we can use `get_shifts()` to easily query the 32,000+ range shifts in the BioShifts database to our target group, target range shift type (here, *latitudinal*), and target continent. 

```{r}

library(BioShiftR)
library(dplyr)

# get shifts for only Birds in Europe
eur_birds <- 
  get_shifts(group = "Birds", # filter to birds
             continent = "Europe", #  filter to Europe
             type = c("LAT","ELE") # latitudinal shifts only
             )  

# view data
eur_birds %>% glimpse()

```


This produces a dataset of `r scales::comma(nrow(eur_birds))` examples of bird range shifts, detected in study areas within the European continent.  

### Assess directional trends

First, we will summarize the proportion of estimated shifts that are in the "expected" directions: First, poleward and upslope. 


```{r, message=F}

library(ggplot2)
eur_birds %>%
  group_by(type) %>%
  summarize(prop_expected = sum(calc_rate > 0)/n()*100) %>%
  ggplot(aes(x = type, y = prop_expected)) +
  geom_col(width = .5) +
  theme_classic() +
  coord_cartesian(xlim = c(.5, 2.5),
                  ylim = c(0,100),
                  expand = F) +
  scale_y_continuous(label = scales::percent_format(scale = 1)) +
  labs(x = NULL, 
       y = "Proportion",
       title = "Shifts going polewards or upslope") +
  theme(plot.title.position = "plot")
  


```


However, sometimes climate expectations can result in isotherm velocities that are not in the upslope or poleward directions. With `BioShiftR`, we can add the climate expectation to further resolve expected shift rates.

```{r}

# add climate velocity
eur_birds_cv <- eur_birds %>%
  # add climate velocity
  add_cv() 
  

eur_birds_cv %>%
  # find proprtion of shifts with same sign as cv
  group_by(type) %>%
  summarize(prop_same_sign = sum(sign(cv_temp_mean) == sign(calc_rate)) / n() * 100) %>%
  
  # plot
    ggplot(aes(x = type, y = prop_same_sign)) +
  geom_col(width = .5) +
  theme_classic() +
  coord_cartesian(xlim = c(.5, 2.5),
                  ylim = c(0,100),
                  expand = F) +
  scale_y_continuous(label = scales::percent_format(scale = 1)) +
  labs(x = NULL, 
       y = "Proportion",
       title = "Shifts going same direction as climate") +
  theme(plot.title.position = "plot")

```


### Assess methodological variables

Shifts are calculated with a wide variety of methods that can affect the precision of their estimates, or at least, limit the comparisons between shifts measured in different ways. 

```{r}

# add methods to birds database
eur_birds_cv_methods <- 
  eur_birds_cv %>%
  add_methods()

# plot based on sample type
eur_birds_cv_methods %>%
  ggplot(aes(x = category,
             y = calc_rate)) +
  geom_boxplot(outliers = F) +
  facet_wrap(~type)



```


